---
title: "My REGEX's summary from R Documentation with CTRL + C, X and V"
author: "Sérgio Quadros"
date: "28-01-2015"
output: html_document
---

## regex {base}: my summary from R Documentation.

### Regular Expressions as used in R

The regular expression patterns supported by *grep* and related functions: *grepl, regexpr, gregexpr, sub, gsub,  strsplit.*

A *‘regular expression’*: a pattern that describes a set of strings.  
Two types:

*  Extended regular expressions (the default): fixed = TRUE which can be considered

*  Perl-like regular expressions: perl = TRUE. 

Patterns are described here as they would be printed by cat: (do remember that backslashes need to be doubled when entering R character strings, e.g. from the keyboard). About this "old line\n new line" they are talking? No, metacharacter versus non-metacharacter...

```{r cat}
cat("(")
cat("\\")

# cat("\(")
## Erro: '\('  é uma seqüência de escape não reconhecida na cadeia de caracteres começando com ""\("
cat("a\nb\nc")
cat(paste(letters, 100* 1:26), fill = TRUE, labels = paste0("{", 1:10, "}:"))
```

Do not assume that **long** regular expressions will be accepted: the POSIX standard only requires up to 256 bytes. Maybe a cause for my grep's problems?

### Extended Regular Expressions (ERE)

We have the regular expressions allowed in the default mode of _grep, regexpr, gregexpr, sub, gsub and strsplit._ They use an implementation of the POSIX 1003.2 standard.

They are similar to arithmetic expressions: various operators to combine smaller expressions. The whole expression matches zero or more characters (read ‘character’ as ‘byte’ if useBytes = TRUE).

The *fundamental building blocks* are the regular expressions that match *a single character*. 

Most characters, including all letters and digits, are regular expressions that match themselves. 

Any metacharacter with special meaning may be quoted by preceding it with a backslash. The metacharacters in EREs are: **. \ | ( ) [ { ^ $ * + ?**, but note that whether these have _a special meaning depends on the context_.

Escaping non-metacharacters with a backslash is implementation-dependent. The current implementation interprets \\a as BEL, \\e as ESC, \\f as FF, \\n as LF, \\r as CR and \\t as TAB. (Note that these will be interpreted by R's parser in literal character strings.)

Metacharacter: 

*  _Character class_: [*list of characters*] which matches _any single character in that list_

*  The caret ^_character_: which matches at first

*  The dollar $ in _character_$: which matches at last

*  _First char_ - _Last char_: hyphen in range of chars: [A-z]=[ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz] Their interpretation is locale; [2-7]; usw.

```{r some grep list}
a <- c("123","234","345","456")
grep("^[2-3]",a)
grep("^[5-8]",a)
grep("[1-3]$",a)
aa <- c("1>23","2c.34","34b:5","45a/6")
aa
gsub("[[:alnum:]]","",aa)
gsub("[:alnum:]","",aa)
gsub("[[:digit:]]","",aa)
gsub("[:digit:]","",aa)
gsub("[[:punct:]]","",aa)
gsub("[:punct:]","",aa)
gsub("[[:alpha:]]","",aa)
gsub("[:alpha:]","",aa)
gsub("[[a:z]]","",aa)
gsub("[a:z]","",aa)
Sys.setlocale()# key hint
```

*  Certain *named classes* of characters are predefined. Their interpretation depends on the locale (see locales); the interpretation below is that of the POSIX locale:


     *  Alphanumeric characters: [:alnum:] = [:alpha:] and [:digit:]; [[:alnum:]] that is independent of locale and character encoding, otherwise [0-9A-Za-z] depends upon the locale and the character encoding. They worked with double square brackets  well.

     *  Alphabetic characters: [:alpha:]

     *  Digits: 0 1 2 3 4 5 6 7 8 9 = [:digit:]

     *  Blank characters: space and tab, and possibly other locale-dependent characters such as non-breaking space = [:blank:]

     *  Lower and uppercase letters in current locale: [:lower:] and [:upper:]


     *  Control characters. In ASCII, these characters have octal codes 000 through 037, and 177 (DEL). In another character set, these are the equivalent characters, if any: [:cntrl:]

     *  Punctuation characters: ! " # $ % & ' ( ) * + , - . / : ; < = > ? @ [ \ ] ^ _ ` { | } ~ = [:punct:]

     *  Graphical characters: [:graph:] = [:alnum:] and [:punct:]

     *  Printable characters: [:print:] = [:alnum:]+[:punct:]+space

     *  Space characters: [:space:] = tab, newline, vertical tab, form feed, carriage return, space and possibly other locale-dependent characters.

     *  Hexadecimal digits: 0 1 2 3 4 5 6 7 8 9 A B C D E F a b c d e f = [:xdigit:]

Most metacharacters lose their special meaning inside a character class. To include a literal ], place it first in the list. Similarly, to include a literal ^, place it anywhere but first. Finally, to include a literal -, place it first or last (or, for perl = TRUE only, precede it by a backslash). (Only ^ - \ ] are special inside character classes.)

*  The period . matches any single character.

*  The symbol \\w matches a ‘word’ character (a synonym for [[:alnum:]_]) and \\W is its negation. 

*  Symbols \\d, \\s, \\D and \\S denote the digit and space classes and their negations.

*  The caret ^ and the dollar sign $ are metacharacters that respectively match the empty string at the beginning and end of a line. 

*  The symbols \\< and \\> match the empty string at the beginning and end of a word.

*  The symbol \\b matches the empty string at either edge of a word, and \\B matches the empty string provided it is not at an edge of a word. 

_The interpretation of ‘word’ depends on the locale and implementation._

*  The repetition quantifiers:

     *  *?* is optional and will be matched at most once.

     *  \* will be matched zero or more times.

     *  \+ will be matched one or more times.

     *  {n} is matched exactly n times.

     *  {n,} is matched n or more times.

     *  {n,m} is matched at least n times, but not more than m times.

By default repetition is greedy, so the maximal possible number of repeats is used. This can be changed to ‘minimal’ by appending ? to the quantifier. (There are further quantifiers that allow approximate matching: see the TRE documentation.)

Regular expressions may be concatenated; the resulting regular expression matches any string formed by concatenating the substrings that match the concatenated subexpressions.

Two regular expressions may be joined by the infix operator |; the resulting regular expression matches any string matching either subexpression. For example, abba|cde matches either the string abba or the string cde. Note that alternation does not work inside character classes, where | has its literal meaning.

Repetition takes precedence over concatenation, which in turn takes precedence over alternation. A whole subexpression may be enclosed in parentheses to override these precedence rules.

The backreference \N, where N = 1 ... 9, matches the substring previously matched by the Nth parenthesized subexpression of the regular expression. (This is an extension for extended regular expressions: POSIX defines them only for basic ones.)

### PCRE - Perl Compatible Regular Expressions

The perl = TRUE argument to grep, regexpr, gregexpr, sub, gsub and strsplit switches to the PCRE library that implements regular expression pattern matching using the same syntax and semantics as Perl 5.10, with just a few differences. *Not yet*

For complete details please consult the man pages for PCRE, especially man pcrepattern and man pcreapi), on your system or from the sources at http://www.pcre.org. 

In UTF-8 mode the named character classes only match ASCII characters: see \p below for an alternative.

regexpr and gregexpr support ‘named capture’. If groups are named, e.g., "(?<first>[A-Z][a-z]+)" then the positions of the matches are also returned by name. (Named backreferences are not supported by sub.)

Atomic grouping, possessive qualifiers and conditional and recursive patterns are not covered here.

#### *Author(s)->from R Documentation*

This help page is based on the documentation of GNU grep 2.4.2, the TRE documentation and the POSIX standard, and the pcrepattern man page from PCRE 8.0.

*See Also*

*grep, apropos, browseEnv, glob2rx, help.search, list.files, ls and strsplit.*

*GREAT*: The TRE documentation at http://laurikari.net/tre/documentation/regex-syntax/).

The POSIX 1003.2 standard at http://pubs.opengroup.org/onlinepubs/9699919799/basedefs/V1_chap09.html#tag_09

The pcrepattern can be found as part of http://www.pcre.org/pcre.txt, and details of Perl's own implementation at http://perldoc.perl.org/perlre.html.

